Differential invariant signatures for planar Lie group transformations of images

### 2021-02-16 paper/newpaper.tex
Updates

\begin{document}
\maketitle
\begin{abstract}
Changes in viewpoint, such as camera motion, impose transformations on the objects in an image, and these transformations will differ for objects at different positions relativele to the camera. Typically these transformations will be elements of planar Lie groups. While there have been significant advances in object recognition using machine learning in recent years, the question of how to recognise an object in an image as its appearance varies through camera motion and similar effects remains open. We consider how differential invariant signatures can be derived for each of the transformation groups, and how the underlying invariances reflect the group-subgroup structure of the transformation groups.

There are a variety of methods that can be used to identify differential invariants, and we provide examples of three of them: tensor contraction, transvectants, and the method of moving frames. We use the resulting invariants to construct practical sets to form three-dimensional invariant signatures. These signatures are not necessarily complete: the image cannot always be reconstructed uniquely up to transformation, but they are plottable, and depend at worst on third derivatives, although more channels of information, such as colour images, can reduce the highest order of derivative needed in some cases. We demonstrate the invariant signatures for each transformation group based on a simple smooth image. A full consideration of how these signatures could be used in practice will require effective methods to numerically approximate derivatives effectively for images.
\end{abstract}

@@

% REQUIRED
\begin{AMS}

\end{AMS}

\section{Introduction}
The appearance of a fixed object in a set of images can vary markedly depending upon such variables as pose and position relative to the camera, even though the object itself does not change. This is even more marked in scenes that consist of multiple objects at different distances and orientations to the camera and any fixed background. The appearance of each individual object will then vary independently as a constant camera motion or other such transformation is applied. Typically the individual transformations will appear as the action of an element of a planar Lie group, for example the affine, or projective groups, or an angle-preserving transformation from the conformal group, depending upon the camera model.

For image recognition it is then natural to seek representations of objects that do not change as these various transformation groups act. If the space of objects is denoted $M$ and the appearance transformations form a group $\mathcal{G}$, then mathematically one seeks a way to study objects in $M / \mathcal{G}$. One way to achieve this is by transforming all objects to match a chosen target object by registration (the transformation of the appearance of one shape to match another, typically by gradient descent on a matching function, see e.g.~\cite{Modersitzki03}). Hence for the group of similarity transformations (rotation and reflection, translation, and scale), a Procrustes alignment~\cite{Kendall1989} will place all of the shapes into the reference frame of one template example. However, analysis on these quotient spaces is rarely simple (see e.g.,~\cite{RiemannianStats} for a basic primer), and if one is interested in the object, and not the transformation, then it necessitates a lot of computational work for relatively little benefit.

An alternative option is to seek $\mathcal{G}$-invariant functions on $M$. Mathematical invariants provide a useful way to identify sets of objects that are the same up to some equivalence relation, since the functions are constant for objects in the same equivalence class. For a group $\mathcal{G}$, orbits of the group form the equivalence classes. For images, rather than considering the image of the object, one constructs an invariant signature of some kind, which represents the object modulo the action of elements from the relevant group. 
This has been considered extensively by Olver and co-authors for the case of outlines of objects~\cite{Calabi1998,Hoff2013}. In this paper we consider the case of signatures for images rather than outlines of curves under a variety of transformation groups.

A greyscale image can be viewed as a two-dimensional manifold embedded in a
three dimensional ambient space, with coordinates $(x, y, f(x,y))$. A
signature has coordinates $\mathcal{I}(f) = (I_1, I_2, \ldots, I_m)$ which
are functions of the derivatives of $f(x, y))$, and will therefore be a
two-dimensional manifold embedded in $\mathbb{R}^m$. Under a change of
coordinates $(x, y) \mapsto (\bar{x}, \bar{y})$ induced by a group action,
the signature of the corresponding image $\bar{f}(\bar{x}, \bar{y}) = f(x, y)$
will have the same signature.

The introduction of an invariant signature reduces the recognition problem to
one of comparing the signatures in a higher-dimensional space that encode the
desired self-equivalences. For greyscale images, it is natural to seek
signatures with three coordinates as $\mathbb{R}^3$ is the smallest space in
which the signature surfaces can reside. In this paper we introduce the
mathematical considerations necessary to construct invariant signatures to
the planar Lie groups (and some infinite-dimensional, pseudo-Lie groups too)
that can transform 2D images. As we shall show, the lattice structure of the
set of groups influences the relationship between the sets of invariants for
the different groups. We provide sample signature sets for the various groups
we consider, and examples of the appearance of those signatures for
transformations of a simple, smooth image, but do not consider the question
of a complete practical implementation. We do not consider the classification
of the signatures, for which either a standard image recognition algorithm
can be used (see, e.g.,~\cite{Zhang20} for a survey), or a matching method
such as~\cite{UsCurrents} can be used.

In order for the invariant signatures to be useful to identify similarity in images, there are two key considerations: they must vary continuously with respect to the image, either for all smooth images or for all except a set of sufficiently high codimension, and they must be robust with respect to occlusions, where one object obscures some part of another. These constraints are essential if the signature is to be used to detect similarity of images, but the first fails for the rational invariants found in the literature. Such considerations lead one naturally to consider local invariants, i.e., those defined using each point of the object as its own basepoint, together with a neighbourhood in some topology. This is sufficient to compute, e.g., derivatives of an object. For this reason we focus on differential invariants. The disadvantage of such invariants, as we shall discuss, is the need to compute several orders of spatial derivative of the image.

Ideally, an invariant signature would be complete, so that the image is exactly determined by the signature up to a transformation, and two objects share a signature if and only if they differ only by a transformation from $\mathcal{G}$. However there are some pathological scenarios that occur regularly in images, particularly that flat regions of images collapse down to single points in the signature as all of the derivatives are zero in these regions. So images that are otherwise similar apart from the size of one of these regions cannot be distinguished between. For this reason, we do not pursue complete signatures, for many applications an incomplete signature is sufficient, and is what we consider here. See, e.g.,~\cite{UsMobius} and references therein for more discussion about the desirable properties of a set of invariants for image and curve analysis.

In this paper we investigate differential invariant signatures for images and parts of images under planar spatial transformations, primarily Lie groups, for both greyscale and colour images. The set of groups that we consider are shown as a lattice in Figure~\ref{fig:lattice}, where the size of the groups decrease as down the page, and the arrows denote the subgroup relationship. There is an important relationship between a group and its subgroups for invariants, which is that invariants of a group are automatically invariants of its subgroups. This means that differentiating between them requires care.

The remainder of this introductory section provides a description of the notation that we use in this paper, followed by a summary of relevant work on invariants. We also introduce the way that we will present examples of the signatures for the various groups we consider. Following this, in the next three sections we describe three distinct methods of computing differential invariants: by tensor contraction, using the theory of transvectants, and via the method of moving frames, applying the methods to a variety of the planar Lie groups. We consider the difference that colour images make in Section~\ref{sec:colour}, and then discuss the number of derivatives required and the effect this has on the practicality of the method.

\begin{figure}
\begin{center}
\begin{tikzpicture}[node distance=2cm]
\node(C1) [] {Diff};
\node(C2) [below left of=C1] {$\mathrm{Diff}_{\mathrm{vol}}$};
\node(C3) [below right of=C1] {$\mathrm{Diff}_{\mathrm{con}}$};
\node(B1)    [below of=C1]    {$PSL(3,\mathbb{R})$};
\node(A1) [below of=C3]                          {$PSL(2,\mathbb{C})$};
\node(B2)   [below of=B1]              {$A(2)$};
\node(A2)       [below right of=B2]          {Sim$(2)$};
%\node(A3)      [right of=A2]              {$PSL(2,\mathbb{R})$};
\node(B3)   [below left of=B2]            {$SA(2)$};
\node(A4)      [below of=A2]            {$E(2)$};
\node(A6)      [below left of=A4]            {$SE(2)$};
\node(A5) [below of=A6]{$T(2)$};

@@

@@

We define a {\em $k$-colour image} as a triple $(f,\Omega,k)$ where $\Omega\subset\R^n$ (typically, images are planar, so $n=2$), $1\le k\in\mathbb{Z}$ is the number of colour channels, and $f\colon\Omega\to\R^k$. The function $f$ will be taken to be as smooth as necessary. Consider a (finite or infinite-dimensional) local transformation group $\mathcal{G}$.  A transformation $\varphi\in G$ acts on images by $\varphi\cdot (f,\Omega,k) = (f\circ\varphi^{-1},\varphi(\Omega),k)$. For many of our examples we consider greyscale images, so $k=1$, although we will discuss colour images in Section~\ref{sec:colour}. Also, since we work locally, we will usually omit the domain $\Omega$.

In order to define differential invariants, it is necessary to identify how the group transformation $\varphi$ acts on derivatives of the function $f$ via the jet space~\cite{OlverEIS}. The $d$-th order jet space $J^d$ consists of the function itself and the set of all partial derivatives of the function up to order $d$. (Note that there are $\left( \begin{array}{c} d+k-1 \\ k \end{array} \right)$ partial derivatives of order $d$.) As the group $\mathcal{G}$ is local, it acts naturally on derivatives of the transformation, and so each individual transformation $\varphi \in \mathcal{G}$ can be prolonged to act on the jet space $J^d$. We can thus prolong the action of the group and consider the action of $\mathcal{G}$ on not just $f$, but all derivatives of $f$ up to order $d$. A $d$-th order differential invariant is then a local scalar function $I: J^{d} (f,\Omega, k) \to \mathbb{R}$ that is invariant under the action of $\mathcal{G}$, i.e., $I(\varphi(f,\Omega, k)) = I(f, \Omega, k) \forall \varphi \in \mathcal{G}$. It can also be useful to consider the maximum order of derivative in the invariant, which we term degree, e.g., $f_{xy}^2 f_x f_y$ has degree 2.

An signature set is the image of $m$ invariants
$\mathcal{I}=(I_1,I_2,\dots,I_m)$. For simplicity of notation we will write
$\mathcal{I}(f)$ with the understanding that each signature component is
actually acting on the jet space $J^d(f, \Omega, k)$. The signature
$\mathcal{I}(f) \subset \mathbb{R}^m$. See Section~\ref{sec:T2} for a basic
example. A signature set is complete for $f$ if it determines $f$ up to
transformations from $\mathcal{G}$, i.e., if $\mathcal{I}(\tilde
f)(\tilde\Omega) = \mathcal{I}(f)(\Omega)$ implies that there exists a
transformation $\varphi\in \mathcal{G}$ such that $\tilde f = \varphi\cdot f$
and $\tilde \Omega = \varphi(\Omega)$. If the signature is not complete, then
there will be a \emph{bad set} of images that cannot be differentiated
between.

The methods we use to construct the invariants, particularly the moving frames, produces invariants that, when considered alone, are singular at certain points (for example, critical points). However, in all groups considered the invariants are rational functions of the derivatives of the function. Moreover, the denominators of these functions are powers of each other, and so it is possible to clear them, thus removing these singular points by a projection technique that will be introduced in Section~\ref{sec:Sim2}.

For the case of two-dimensional greyscale images it is often convenient to work in coordinates. In this case the transformation group action by a coordinate transformation $\varphi \colon \mathbb{R}^2 \to \mathbb{R}^2$ is given by:
\begin{equation}\label{eq:transformation}
  \mathbf{x} \mapsto \bar{\mathbf{x}} = \varphi(\mathbf{x}).
\end{equation}
The transformed image is then defined to be:
\begin{equation}\label{eq:fbarequalsf}
  \bar{f}(\bar{\mathbf{x}}) = (f\circ\varphi^{-1})(\bar{\mathbf{x}}) = f(\mathbf{x}).
\end{equation}

@@

%Groups we've done:
%translations, E(2), Sim(2), SA(2), A(2), PSL(2,C), PSL(3,R), Diff\_con, Diff\_vol, Diff
\subsection{Test image}
In this paper our aim is develop the mathematical underpinnings of
constructing differential invariant signatures for images under the action
of planar Lie groups. However, as we develop the signature for each group,
we will provide an example of the type of surface that the signature
defines. Note that these will generally be three-dimensional representation
of the planar image.

The image we use is defined on $[-1,1] \times [-1,1]$ to be:

@@

The function is depicted as a greyscale image in Figure~\ref{fig:function}.
On the left the image itself is shown, while on the right a version with a
regular grid superimposed is given. We will use this function throughout,
using the grid to help visualise both the transformation and how the image
is mapped into the three-dimensional signature surface.

\begin{figure}
  \centering
  \includegraphics[width=.45\textwidth]{Figs/function}
  \includegraphics[width=.45\textwidth]{Figs/function_scanlines}
  \caption{The test function used in this document, from
  Eq.~\eqref{eq:function}.}\label{fig:function}
\end{figure}

@@

We could explicitly write the calculation in Eq.~\ref{eq:prolongation} for
this case, but instead provide an intuitive explanation. Consider a point
in the original image, and the corresponding point in the transformed one.
Clearly, the intensity value of the points match, and so do their first
spatial derivatives. These three quantities can be used as coordinates in a
new space, where the two image points will correspond. Computing these
coordinates for all image points will define a surface in the new
three-dimensional space, and the two images will produce the same surface,
which we call the signature, $\mathcal{I}(f)(x, y) = \left(I_1, I_2,
I_3\right)$ where $I_1 = f, I_2 = f_x, I_3 = f_y$.

Figure~\ref{fig:T2} shows (a) the original image, (b) the translated image
under the transformation $(x,y) \mapsto (x + 0.1, y - 0.2)$, (c) the
signature of one of them, (d) the superimposed pair of signatures. The
yellow lines in (c) correspond to where the yellow grid from (a) has been
mapped to on the signature surface. The parts of the surface where the mesh
gets close together are where the function and its derivatives are tending
towards zero. This occurs around the boundary of our test image and so all
of these points are close together in the signature space. In (d), notice
that as some of the image has moved out of frame, there are parts of the
signature surfaces of the two images that do not overlap.

\begin{figure}
  \centering
  \begin{tabular}{cc}
  \includegraphics[width=.45\textwidth]{Figs/function_scanlines} &
  \includegraphics[width=.45\textwidth]{Figs/f_transformed_T2} \\
  (a) & (b) \\
  \includegraphics[width=.45\textwidth]{Figs/T2_signature} &
  \includegraphics[width=.45\textwidth]{Figs/T2_match} \\
  (c) & (d)
  \end{tabular}
  \caption{(a), (b) The image before and after the T(2) transformation $(x,
  y) \mapsto (x + 0.1, y - 0.2)$. (c) the signature of the
  image (before translation). (d) the signature after the
  translation superimposed.}\label{fig:T2}
\end{figure}

@@

The theory of invariants has a rich mathematical history, starting from invariants of forms~\cite{Cayley}, where the invariants are polynomials in the coefficients of the forms, see e.g.,~\cite{OlverCIT,Kraft2002} for an overview. Geometric considerations and differential invariants followed fairly quickly~\cite{Veblen1872,Tresse1894,Bouton1898,Patterson1928}, particularly because they enabled the classification of differential equations into equivalence classes~\cite{Lie1884,Littlewood44, Cartan52}. More recently, this has extended, primarily by Olver and collaborators, see e.g.,~\cite{OlverCIT} and~\cite{OlverEIS} for summaries.

There has been a long-standing interest in constructing invariants for image recognition and pattern analysis, see e.g.,~\cite{Mundy1992} and~\cite{Wood1996}. This was set in the context of Lie's invariants in~\cite{VanGool1995}, where the concept of a joint (or semi-) differential invariant (an invariant that acts on the joint action of the transformation group, see~\cite{Fels1997,Olver2013}) was introduced~\cite{VanGool1992}; see also~\cite{Hubert2009}.

A variety of function spaces have allowed for the creation of different sets of invariants for particular groups. For the Euclidean group, the Fourier transform provides a useful tool~\cite{Ghorbel1994,Turski2006,Smach2007,Gauthier2008}, and this can be extended to the affine group using the Fourier-Mellin transform~\cite{Reddy1996,Zhang2019} and the bispectrum~\cite{Kakarala2009,Negrinho2013}. Alternatively, integral invariants~\cite{Manay2006,Feng2010} and geometric moments can be computed. For the similarity group applied to images the original reference is \cite{Hu1962}, but see also~\cite{Hickman2011} for affine transformations, \cite{Flusser2009} for an implicit representation, and \cite{Papakostas2013} for their use in a binary pattern for image recognition, which is somewhat related to the invariant features such as SURF~\cite{SURF}.

The group that has been of particular interest for image analysis is the projective group, because of its ubiquity in camera models. Invariants for this group have been computing using points~\cite{Gros1992,Suk2000}, geometric moments~\cite{Suk2004}, curvature~\cite{Hann2002a}, and differential and joint invariants~\cite{Hubert2007,Arora2009,Kogan2014,Li19}.

Although there are computational challenges regarding the numerical approximation of derivatives on images, the fact that differential invariants are robust to occlusion, cheap to compute, and well-behaved with respect to small changes in the image, has made them popular for researchers in image analysis~\cite{Florack1993}. Two particular tools have been key to this: Cartan's moving frame~\cite{Cartan35} (see Section~\ref{sec:movingframes}), and which was used to construct image invariants as early as 1994 in \cite{Faugeras1994}, and the concept of an image signature built from a set of invariants~\cite{Calabi1998}. An algorithm to compute differential and joint invariants and determine complete signature sets is described in \cite{Olver2005,Olver2007}. The entire set of invariants is computed using the method of moving frames and then the signature set of a maximal set of functionally independent invariants is identified. A third step is possible, in which the dimension of the signature set is reduced by eliminating those invariants whose values can be computed from the values of derivatives of some subset of the invariants.

@@

Another way to compute invariants is to compare the simultaneous action of a group element on separate copies of the underlying space (here, $\mathbb{R}^2$). By identifying these spaces with each other it is possible to construct a function known as a transvectant that is invariant up to a scale factor (a \emph{relative invariant}) to that group action.

>>>>

\begin{document}
\maketitle
\begin{abstract}
Changes in viewpoint, such as camera motion, impose transformations on the objects in an image, and these transformations will differ for objects at different positions relative to the camera. Typically these transformations will be elements of planar Lie groups. While there have been significant advances in object recognition using machine learning in recent years, the question of how to recognise an object in an image as its appearance varies through camera motion and similar effects remains open. We demonstrate how differential invariant signatures can be derived for each of the transformation groups, and how the underlying invariances reflect the group-subgroup structure of the transformation groups.

There are a variety of methods that can be used to identify differential invariants, and we provide examples of three of them: tensor contraction, transvectants, and the method of moving frames. We use the resulting invariants to construct practical sets to form three-dimensional invariant signatures. These signatures are not necessarily complete: the image cannot always be reconstructed uniquely up to transformation, but they are plottable, and depend at worst on third derivatives, although more channels of information, such as colour images, can reduce the highest order of derivative needed in some cases. We demonstrate the invariant signatures for each transformation group based on a simple smooth image. A full consideration of how these signatures could be used in practice will require effective methods to numerically approximate derivatives for images.
\end{abstract}

@@

% REQUIRED
\begin{AMS}
65D18, 53A55, 68U10
\end{AMS}

\section{Introduction}
The appearance of a fixed object in a set of images can vary markedly depending upon such variables as pose and position relative to the camera, even though the object itself does not change. This is even more marked in scenes that consist of multiple objects at different distances and orientations to the camera against a fixed background. The appearance of each individual object will then vary independently as a constant camera motion or other such transformation is applied. Typically the individual transformations will appear as the action of an element of a planar Lie group, for example the affine, or projective groups, or an angle-preserving transformation from the conformal group, depending upon the camera model.

For image recognition it is then natural to seek representations of objects that do not change as these various transformation groups act. If the space of objects is denoted $M$ and the appearance transformations form a group $\mathcal{G}$, then mathematically one seeks a way to study objects in $M / \mathcal{G}$. One way to achieve this is by transforming all objects to match a chosen target object by registration (the transformation of the appearance of one shape to match another, typically by gradient descent on a matching function, see e.g.,~\citep{Modersitzki03}). For example, for the group of similarity transformations (rotation and reflection, translation, and scale), a Procrustes alignment~\citep{Kendall1989} will place all of the shapes into the reference frame of one template example. However, analysis on these quotient spaces is rarely simple (see e.g.,~\citet{RiemannianStats} for a basic primer), and if one is interested in the object, and not the transformation, then it necessitates a lot of computational work for relatively little benefit. The alternative approach, of showing examples of each object after applying sufficiently many transformations that the classifier can learn to recognise the object after any group transformation, inevitably scales extremely badly.

An alternative option is to seek $\mathcal{G}$-invariant functions on $M$. Mathematical invariants provide a useful way to identify sets of objects that are the same up to some equivalence relation, since the functions are constant for objects in the same equivalence class (orbits of the group). For images, rather than considering the image of the object, one constructs an invariant signature of some kind, which represents the object modulo the action of elements from the relevant group. This has been considered extensively by Olver and co-authors for the case of outlines of objects~\citep{Calabi1998,Hoff2013}. We extend this by considering the case of signatures for images rather than outlines of curves under a variety of transformation groups of relevance to image analysis.

A greyscale image can be viewed as a two-dimensional manifold embedded in a three dimensional ambient space, with coordinates $(x, y, f(x,y))$. A signature $\mathcal{I}(f) \in \mathbb{R}^m$ has coordinates $\mathcal{I}(f) = (I_1, I_2, \ldots, I_m)$, where each $I_i$ is a function of $f(x, y))$ and its derivatives, and will therefore be a two-dimensional manifold embedded in $\mathbb{R}^m$. Under a change of coordinates $(x, y) \mapsto (\bar{x}, \bar{y})$ induced by a group action, the signatures of the original image and transformed version $\bar{f}(\bar{x}, \bar{y}) = f(x, y)$ will be the same. The introduction of an invariant signature reduces the object recognition problem modulo some transformation group to one of comparing the signatures of the objects, rather than the images themselves. We do not consider the question of classification further in this paper; a standard image recognition algorithm can be used (see, e.g.,~\citet{Zhang20} for a survey), or a matching method such as~\citet{UsCurrents} can be used.

Ideally, an invariant signature would be complete, so that the image is exactly determined by the signature up to a transformation, and two objects share a signature if and only if they differ only by a transformation from $\mathcal{G}$. However, there are some pathological scenarios that occur regularly in images, particularly that flat regions of images collapse down to single points in the signature (as all of the derivatives are zero in these regions). Then images that are otherwise similar (apart from the size of one of these regions) cannot be distinguished. However, the number of dimensions $m$ of the ambient manifold can increase markedly for complete signatures, increasing the computational cost and making it impossible to visualise the signature surface. For many applications an incomplete signature is sufficient, and is what we consider here, favouring signatures in three coordinates, since for greyscale images, $\mathbb{R}^3$ is the smallest space in which the signature surfaces can reside.  If the signature is not complete, then there will be a `bad set' of images that cannot be distinguished from each other.  The codimension of the bad set is defined as the difference between the dimensionalities of the whole space and of the bad set. The codimension should ideally be large, meaning that the dimension of the bad set would be small.

In addition to a bad set of high codimension, in order for the invariant signatures to be useful to identify similarity in images, there are two other key considerations: they must vary continuously with respect to the image, and they must be robust with respect to occlusions, where one object obscures some part of another. These constraints are essential if the signature is to be used to detect similarity of images, but the first fails for the rational invariants found in the literature. Such considerations lead one naturally to consider local invariants, i.e., those defined using each point of the object as its own basepoint, together with a neighbourhood in some topology. This is sufficient to compute, e.g., derivatives of an object. For this reason we focus on differential invariants. The disadvantage of such invariants, as we shall discuss, is the need to compute several orders of spatial derivative of the image. See~\citet{UsMobius} and references therein for more discussion about the desirable properties of a set of invariants for image and curve analysis.

In this paper our aim is develop the mathematical underpinnings of constructing differential invariant signatures for images under the action of planar Lie groups, and to summarise the impediments to a practical implementation, including the number of derivatives required. We demonstrate three distinct methods of computing differential invariants: by tensor contraction, using the theory of transvectants, and via the method of moving frames, applying the methods to a variety of the planar Lie groups, providing sample signature sets for the various groups we consider, together with an example of a sample signature surface based on a random transformation of a simple, smooth image.

The set of groups that we consider are shown as a lattice in Fig.~\ref{fig:lattice}, where the size of the groups decrease down the page, and the arrows denote the subgroup relationship.  There is an important relationship between a group and its subgroups for invariants, which is that invariants of a group are automatically invariants of its subgroups. This means that care is needed to distinguish between the two, with further invariants that are known not to be invariant in the supergroup included. The finite-dimensional groups in this lattice were chosen because they are the only planar Lie groups that include translation and have locally primitive algebras (that is, for any open subset of the plane, there is no one-dimensional foliation that is left invariant by the action of the corresponding group~\citep{Gonzalez-Lopez90}). This makes them the groups most directly relevant to image transformations. The infinite-dimensional examples are likewise those most likely to be of interest for image transformation, in particular the full diffeomorphism group, which is commonly used for image registration for medical images, see e.g.,~\citet{YounesBook}.

%We next provide a description of the notation that we use in this paper, followed by a summary of relevant work on invariants. We also introduce the way that we will present examples of the signatures for the various groups we consider. Following this, in the next three sections we describe three distinct methods of computing differential invariants: by tensor contraction, using the theory of transvectants, and via the method of moving frames, applying the methods to a variety of the planar Lie groups. We conclude by considering the practical issues of implementation, including the number of derivatives required and the effect this has on the practicality of the method.

\begin{figure}
\begin{center}
\begin{tikzpicture}[node distance=2cm]
\node(C1) [] {$\mathrm{Diff}$};
\node(C2) [below left of=C1] {$\mathrm{Diff}_{\mathrm{vol}}$};
\node(C3) [below right of=C1] {$\mathrm{Diff}_{\mathrm{con}}$};
\node(B1)    [below of=C1]    {$PSL(3,\mathbb{R})$};
\node(A1) [below of=C3]                          {$PSL(2,\mathbb{C})$};
\node(B2)   [below of=B1]              {$A(2)$};
\node(A2)       [below right of=B2]          {Sim$(2)$};
%\node(A3)      [right of=A2]              {$PSL(2,\mathbb{R})$};
\node(B3)   [below left of=B2]            {$SA(2)$};
\node(A4)      [below of=A2]            {$E(2)$};
\node(A6)      [below left of=A4]            {$SE(2)$};
\node(A5) [below of=A6]{$T(2)$};

@@

\section{Preliminaries}

@@

We define a {\em $k$-colour image} as a triple $(f,\Omega,k)$ where $\Omega\subset\R^n$ (typically, images are planar, so $n=2$), $1\le k\in\mathbb{Z}$ is the number of colour channels, and $f\colon\Omega\to\R^k$. The function $f$ will be taken to be as smooth as necessary. Consider a (finite or infinite-dimensional) local transformation group $\mathcal{G}$.  A transformation $\varphi\in G$ acts on images by $\varphi\cdot (f,\Omega,k) = (f\circ\varphi^{-1},\varphi(\Omega),k)$. We mostly consider greyscale images, so $k=1$. Also, since we work locally, we will usually omit the domain $\Omega$.

In order to define differential invariants, it is necessary to identify how the group transformation $\varphi$ acts on derivatives of the function $f$ via the jet space~\citep{OlverEIS}. The $d$-th order jet space $J^d$ consists of the function itself and the set of all partial derivatives of the function up to order $d$. (Note that there are $\left( \begin{array}{c} d+k-1 \\ k \end{array} \right)$ partial derivatives of order $d$.) As the group $\mathcal{G}$ is local, it acts naturally on derivatives of the transformation, and so each individual transformation $\varphi \in \mathcal{G}$ can be prolonged to act on the jet space $J^d$. We can thus prolong the action of the group and consider the action of $\mathcal{G}$ on not just $f$, but all derivatives of $f$ up to order $d$. A $d$-th order differential invariant is then a local scalar function $I: J^{d} (f,\Omega, k) \to \mathbb{R}$ that is invariant under the action of $\mathcal{G}$, i.e., $I(\varphi(f,\Omega, k)) = I(f, \Omega, k) \forall \varphi \in \mathcal{G}$. It can also be useful to consider the maximum order of derivative in the invariant, which we term degree, e.g., $f_{xy}^2 f_x f_y$ has degree 2.

An signature set is the image of $m$ invariants
$\mathcal{I}=(I_1,I_2,\dots,I_m)$. For simplicity of notation we will write
$\mathcal{I}(f) \subset \mathbb{R}^m$ with the understanding that each signature component is
actually acting on the jet space $J^d(f, \Omega, k)$. See Section~\ref{sec:T2} for a basic
example. A signature set is complete for $f$ if it determines $f$ up to
transformations from $\mathcal{G}$, i.e., if $\mathcal{I}(\tilde
f)(\tilde\Omega) = \mathcal{I}(f)(\Omega)$ implies that there exists a
transformation $\varphi\in \mathcal{G}$ such that $\tilde f = \varphi\cdot f$
and $\tilde \Omega = \varphi(\Omega)$.

The methods we use to construct the invariants, particularly the moving frames, produces invariants that, when considered alone, are singular at points (such as critical points). However, in all groups considered the invariants are rational functions of the derivatives of the function. Moreover, the denominators of these functions are powers of each other, and so it is possible to clear them by a projection technique that will be introduced in Section~\ref{sec:Sim2}.

For the case of two-dimensional greyscale images it is often convenient to work in coordinates. In this case the transformation group action by a coordinate transformation $\varphi \colon \mathbb{R}^2 \to \mathbb{R}^2$ is given by (where $f$ and $\mathbf{x}$ denote the original image and coordinates, and $\bar{f}$ and $\bar{\mathbf{x}}$ the transformed versions):
\begin{equation}\label{eq:transformation}
  \mathbf{x} \mapsto \bar{\mathbf{x}} = \varphi(\mathbf{x}).
\end{equation}
The transformed image is then defined to be:
\begin{equation}\label{eq:fbarequalsf}
  \bar{f}(\bar{\mathbf{x}}) = (f\circ\varphi^{-1})(\bar{\mathbf{x}}) = f(\mathbf{x}).
\end{equation}

@@

%Groups we've done:
%translations, E(2), Sim(2), SA(2), A(2), PSL(2,C), PSL(3,R), Diff\_con, Diff\_vol, Diff
\subsection{Test image}

For each signature set that we derive, it is informative to see an example signature of the kind of surface that is defined, 
normally a two-dimensional surface in three-dimensions.  We will apply a small random transformation from 
the relevant group to a single continuous image and plot the signature before and after this transformation has been applied.
This is an aid to understanding rather than any form of demonstration of the robustness of our signature sets.

The image we use is a continuous function (thus avoiding numerical issues in the approximation 
of derivatives) defined on $[-1,1] \times [-1,1]$ as:

@@

The function is depicted as a greyscale image in Fig.~\ref{fig:function}.
We will use this function throughout, using the grid on the right of the figure to help visualise both the transformation and how the image
is mapped into the three-dimensional signature surface.

\begin{figure}
  \centering
  \includegraphics[width=.45\textwidth]{Figs/function}
  \includegraphics[width=.45\textwidth]{Figs/function_scanlines}
  \caption{The test function used as an image, defined in Eq.~\eqref{eq:function}.
On the left the image itself is shown, while on the right a version with a
regular grid superimposed is given. 
}\label{fig:function}
\end{figure}

@@

We could explicitly write the calculation in Eq.~\ref{eq:prolongation} for
this case, but instead provide an intuitive explanation. Consider a point
in the original image, and the corresponding point in the transformed one.
Clearly, the intensity value of the points match, and so do their first
spatial derivatives. These three quantities can be used as coordinates in a
new space, where the two image points will correspond. Computing these
coordinates for all image points will define a surface in the new
three-dimensional space, and the two images will produce the same surface,
the signature, $\mathcal{I}(f)(x, y) = \left(I_1, I_2,
I_3\right)$ where $I_1 = f, I_2 = f_x, I_3 = f_y$.

Fig.~\ref{fig:T2} shows (a) the original image, (b) the translated image
under the transformation $(x,y) \mapsto (x + 0.1, y - 0.2)$, (c) the
signature of one of them, (d) the superimposed pair of signatures. The
yellow lines in (c) correspond to places where the yellow grid from (a) has been
mapped to on the signature surface. The parts of the surface where the mesh
gets close together are where the function and its derivatives are tending
towards zero. This occurs around the boundary of our test image and so all
of these points are close together in the signature space. In (d), notice
that as some of the image has moved out of frame, there are parts of the
signature surfaces of the two images that do not overlap.

\begin{figure}
  \centering
  \begin{tabular}{cc}
  \includegraphics[width=.45\textwidth]{Figs/function_scanlines} &
  \includegraphics[width=.45\textwidth]{Figs/f_transformed_T2} \\
  (a) & (b) \\
  \includegraphics[width=.45\textwidth]{Figs/T2_signature} &
  \includegraphics[width=.45\textwidth]{Figs/T2_match} \\
  (c) & (d)
  \end{tabular}
  \caption{(a), (b) The image before and after the $T(2)$ transformation $(x,
  y) \mapsto (x + 0.1, y - 0.2)$. (c) the signature of the
  image (before translation). (d) the signatures before and after the
  translation superimposed.}\label{fig:T2}
\end{figure}

@@

The theory of invariants has a rich mathematical history, starting from invariants of forms~\citep{Cayley}, where the invariants are polynomials in the coefficients of the forms, see e.g.,~\citet{OlverCIT,Kraft2002} for an overview. Geometric considerations and differential invariants followed fairly quickly~\citep{Veblen1872,Tresse1894,Bouton1898,Patterson1928}, particularly because they enabled the classification of differential equations into equivalence classes~\citep{Lie1884,Littlewood44, Cartan52}. More recently, this has been extended, primarily by Olver and collaborators, see e.g.,~\citet{OlverCIT} and~\citet{OlverEIS} for summaries.

There has been a long-standing interest in constructing invariants for image recognition and pattern analysis, see e.g.,~\citet{Mundy1992} and~\citet{Wood1996}. This was set in the context of Lie's invariants by~\citet{VanGool1995}, where the concept of a joint (or semi-) differential invariant (an invariant that acts on the joint action of the transformation group, see~\citet{Fels1997,Olver2013}) was introduced~\citep{VanGool1992}; see also~\citet{Hubert2009}.

A variety of function spaces have allowed for the creation of different sets of invariants for particular groups. For the Euclidean group, the Fourier transform provides a useful tool~\citep{Ghorbel1994,Turski2006,Smach2007,Gauthier2008}, and this can be extended to the affine group using the Fourier-Mellin transform~\citep{Reddy1996,Zhang2019} and the bispectrum~\citep{Kakarala2009,Negrinho2013}. Alternatively, integral invariants~\citep{Manay2006,Feng2010} and geometric moments can be computed. For the similarity group applied to images the original reference is \citet{Hu1962}, but see also~\citet{Hickman2011} for affine transformations, \citet{Flusser2009} for an implicit representation, and \citet{Papakostas2013} for their use in a binary pattern for image recognition, which is somewhat related to the invariant features such as SURF~\citep{SURF}.

The group that has been of particular interest for image analysis is the projective group, because of its ubiquity in camera models. Invariants for this group have been computing using points~\citep{Gros1992,Suk2000}, geometric moments~\citep{Suk2004}, curvature~\citep{Hann2002a}, and differential and joint invariants~\citep{Hubert2007,Arora2009,Kogan2014,Li19}.

Although there are computational challenges regarding the numerical approximation of derivatives on images, the fact that differential invariants are robust to occlusion, cheap to compute, and well-behaved with respect to small changes in the image, has made them popular for researchers in image analysis~\citep{Florack1993}. Two particular tools have been key to this: Cartan's moving frame~\citep{Cartan35} (see Section~\ref{sec:movingframes}), and which was used to construct image invariants as early as 1994 by \citet{Faugeras1994}, and the concept of an image signature built from a set of invariants~\citep{Calabi1998}. An algorithm to compute differential and joint invariants and determine complete signature sets is described by \citet{Olver2005,Olver2007}. The entire set of invariants is computed using the method of moving frames and then the signature set of a maximal set of functionally independent invariants is identified. A third step is possible, in which the dimension of the signature set is reduced by eliminating those invariants whose values can be computed from the values of derivatives of some subset of the invariants.

@@

Another way to compute invariants is to compare the simultaneous action of a group element on separate copies of the underlying space (for planar images, $\mathbb{R}^2$). By identifying these spaces with each other it is possible to construct a function known as a transvectant that is a relative invariant to that group action.
